QuickSched: Task-based parallelism with dependencies and conflicts

### 2015-06-09 paper/paper.tex
started porting to PeerJ template.

\documentclass[preprint]{elsarticle}

@@

@@

% Title portion
\title{QuickSched: Task-based parallelism with dependencies and conflicts}
\author[ecs,ggl]{Pedro Gonnet\corref{cor}} \ead{pedro.gonnet@durham.ac.uk}
\author[ecs]{Aidan B. G. Chalk} \ead{aidan.chalk@durham.ac.uk}
\author[icc]{Matthieu Schaller} \ead{matthieu.schaller@durham.ac.uk}
\cortext[cor]{Corresponding author}
\address[ecs]{School of Engineering and Computing Sciences, Durham University, South Road, Durham DH1 3LE, United Kingdom.}
\address[icc]{Institute for Computational Cosmology, Durham University, South Road, Durham DH1 3LE, United Kingdom.}
\address[ggl]{Google Switzerland GmbH, Brandshenkestr. 110, 8002 Z\"urich, Switzerland.}

\begin{keyword}
    task-based parallelism \sep shared-memory parallelism \sep high performance computing \sep parallel computing
\end{keyword}

\maketitle

@@

This computational model is trivial to parallelize.
Given a set of inter-dependent tasks and a set of computational
threads, each thread repeatedly selects a task with no
unsatisfied dependencies from the DAG and executes it.
If no tasks are available, the thread waits until any other
thread finishes executing a task, thus potentially releasing
new tasks, or until all tasks in the DAG have been executed.
\fig{Tasks} shows such a DAG for a set of tasks.
The arrows indicate the direction of the dependency, i.e.~an
arrow from task $A$ to task $B$ indicates that task $B$ depends
on task $A$.
In a parallel setting, tasks $A$, $G$, and $J$ can be
executed concurrently.
Once task $G$ has completed, tasks $F$ and $H$ become available
and can be executed by any other computational thread.

@@

One of the first implementations of a task-based parallel programming
systems is Cilk \cite{ref:Blumofe1995}, an extension to the C
programming language which allows function calls to be ``spawned''
as new tasks.
Dependencies are enforced by the {\tt sync} keyword, which
forces a thread to wait for all the tasks that it spawned
to complete.

In SMP superscalar \cite{ref:Perez2008}, StarPU \cite{ref:Augonnet2011},
QUARK \cite{ref:Yarkhan2011}, and KAAPI \cite{ref:Gautier2007}
the programmer specifies
what shared data each task will access, and how that data will
be accessed, e.g.~read, write, or read-write access.
The dependencies between tasks are then generated
automatically by the runtime system, assuming that the
data must be accessed and updated in the order in which
the tasks are generated.
StarPU also provides an interface for specifying additional
dependencies explicitly.
Intel's Threading Building Blocks (TBB)
\cite{ref:Reinders2010}
provide task-based parallelism using C++ templates in which
dependencies are handled either by explicitly waiting
for spawned tasks, or by explicitly manipulating
task reference counters.

Finally, the very popular OpenMP standard provides some basic support
for spawning tasks, similar to Cilk, as of version 3.0
\cite{ref:OpenMP2008}.
OmpSs \cite{ref:Duran2011} extends this scheme with automatic
dependency generation as in SMP superscalar, of which it
is a direct descendant, along with
the ability to explicitly wait on certain tasks.

In all of these systems, the tasks are only aware of a single
type of relationship between each other, i.e. dependencies, which
specify a strict ordering between two tasks.
In many cases, however, the task ordering need not necessarily
be this strict.
Consider the case of two tasks that update some shared resource
in an order-independent way, e.g. when accumulating a result in
a shared variable, or exclusively writing to an output file.
In order to avoid concurrent access to that resource, it is
imperative that the execution of both tasks do not overlap,
yet the order in which the tasks are executed is irrelevant.
In the following, such a relationship will be referred to
as a {\em conflict} between two tasks.
\fig{TaskConflicts} shows a task graph extended by conflicting tasks
joined by thick dashed lines.
None of tasks $F$, $H$, and $I$ can be executed concurrently,
i.e. they must be serialized, yet in no particular order.

@@

From a programmer's perspective, there are two main paradigms for generating
task dependencies:
\begin{itemize}
  \item Implicitly via spawning and waiting, e.g.~as is done in Cilk
    \cite{ref:Blumofe1995}, or
  \item Automatic extraction from data dependencies, e.g.~as is done in OmpSs
    \cite{ref:Duran2011}.
\end{itemize}

@@

In the examples presented herein, we have chosen our task decomposition
and granularity such that
\begin{itemize}
  \item Each task maximizes the ratio of computation to data required,
  \item The resources required for each task fit comfortably in the
    lowest-level caches of the underlying system.
\end{itemize}
\noindent The first critera is biased towards bigger tasks, while the
second limits their size.
The criteria are thus used to optimize the
cache efficiency of the computation.

@@

The QuickSched task scheduler consist of four main
objects types: {\em task}, {\em resource}, {\em scheduler},
and {\em queue}.
The task and resource objects are used
to model the computation, i.e. the work that is to be done
and the data on which it will be done, respectively.
The scheduler and queue objects manage
how the work is done, i.e. which tasks get scheduled
where and when, respectively.

@@

Finally, {\tt cost} and {\tt weight} are measures
for the relative computational cost of this task, and the
relative cost of the critical path following the
dependencies of this task, respectively, i.e.~the task's cost
plus the maximum dependent task's weight (see \fig{TaskWeight}).
The task weights are computed by traversing
the tasks in reverse topological order following their dependencies, i.e.
as per \cite{ref:Kahn1962}, and computing each task's weight
up the task DAG.

@@

@@

The tasks are executed as follows:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
void qsched_run(qsched *s, void (*fun)(int, void *)) {
  qsched_start(s);
  #pragma omp parallel
  {
    int qid = omp_get_thread_num() % s->nr_queues;
    struct task *t;
    while ((t = qsched_gettask(s, qid)) != NULL) {
      fun(t->type, t->data);
      qsched_done(s, t);
    }
  }
}
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where {\tt qsched\_start} initializes the tasks and
fills the queues (line~1).
For simplicity, OpenMP \cite{ref:Dagum1998}, which is available
for most compilers, is used to create a parallel section
in which the code between lines~4 and~11 is executed
concurrently.
A version using {\tt pthreads} \cite{ref:Pthreads1995}
directly\footnote{In most environments, OpenMP is implemented
on top of {\tt pthreads}, e.g. gcc's libgomp.} is also available.
The parallel section consists of a loop (lines~7--10) in
which a task is acquired via {\tt qsched\_gettask}
and its type and data are passed to a user-supplied
{\em execution function} {\tt fun}.
Once the task has been executed, it is returned to the
scheduler via the function {\tt qsched\_done}, i.e.~to
unlock its resources and unlock dependent tasks.
The loop terminates when the scheduler runs out of tasks,
i.e.~when {\tt qsched\_gettask} returns {\tt NULL}, and
the function exits once all the threads have exited their
loops.

@@

The function {\tt qsched\_gettask} fetches a task from
one of the queues:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
struct task *qsched_gettask(qsched *s, int qid) {
  struct task *res = NULL;
  int k;
  while (s->waiting) {
    if ((res = queue_get(s->queues[qid])) == NULL) {
      loop over all other queues in random order with index k
        if ((res = queue_get(s->queues[k])) != NULL)
          break;
    } else
      break;
  }
  if (res != NULL && s->reown) {
    for (k = 0; k < res->nr_locks; k++)
      res->locks[k]->owner = qid;
    for (k = 0; k < res->nr_uses; k++)
      res->uses[k]->owner = qid;
  }
  return res;
}
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where the parameter {\tt qid} is the index of the
preferred queue.
If the queue is empty, or all of the tasks in that queue had
unresolved conflicts, the scheduler uses a variant of the
{\em work stealing} described in \cite{ref:Blumofe1999},
i.e.~it loops over all other queues
in a random order (line~6) and tries to get a task from them
(line~7).
If a task could be obtained from any queue and task re-owning
is switched on (line~12),
the resources it locks and uses are marked as now being owned
by the preferred queue (lines~13--16).
Finally, the task, or {\tt NULL} if no task could be obtained,
is returned.

@@

This section presents two test cases that show
how QuickSched can be used in real-world applications, and
provides benchmarks to assess its efficiency and scalability.
The first test is the tiled QR decomposition originally
described in \cite{ref:Buttari2009}, which has been used as a benchmark
by other authors \cite{ref:Agullo2009b,ref:Badia2009,ref:Bosilca2012}.
This example only requires dependencies and is presented 
as a point of comparison to existing task-based parallel
programming infrastructures.
The second example is a Barnes-Hut tree-code, a problem
similar to the Fast Multipole Method described in both
\cite{ref:Ltaief2012} and \cite{ref:Agullo2013}.
This example shows how conflicts, modeled
via hierarchical resources, can be useful in modelling and executing
a problem efficiently.

@@

\cite{ref:Buttari2009} introduced the concept of using task-based
parallelism for tile-based algorithms in numerical linear algebra,
presenting parallel codes for the Cholesky, LU, and QR
decompositions.
These algorithms are now part of the PLASMA and MAGMA
libraries for parallel linear algebra \cite{ref:Agullo2009}.
The former uses the QUARK task scheduler, which was originally
designed for this specific task, while the latter currently uses
the StarPU task scheduler \cite{ref:Agullo2011}.

@@

\noindent and where the task names are the BLAS-like operation
performed on the given tiles.
Every task depends on the task at the same position and the
previous level, i.e.~the task $(i,j,k)$ always depends on
$(i,j,k-1)$ for $k>1$.
Each task also modifies its own tile $(i,j)$, and the DTSQRF
task additionally modifies the lower triangular part of the $(j,j)$th tile.
Although the tile-based QR decomposition requires only dependencies,
i.e.~no additional conflicts are needed to avoid concurrent access to
the matrix tiles, we still model each tile as a separate resource
in QuickSched such that the scheduler can preferrentially assign
tasks using the same tiles to the same thread.

The QR decomposition was computed for a $2048\times 2048$
random matrix using tiles of size $64\times 64$ floats using QuickSched
as described above.
For this matrix, a total of 11440 tasks with 32240 dependencies
were generated.

For these tests, {\tt pthread} parallelism and resource re-owning
were used with one queue per core.
The QR decomposition was computed 10 times for each number of
cores, and the average thereof taken for the scaling and
efficiency results in \fig{QRResults}.
The timings are for {\tt qsched\_run}, including the cost of
{\tt qsched\_start}, which does not run in parallel.
Setting up the scheduler, tasks, and resources took, in all
cases, an average of 7.2\,ms.

The same decomposition was implemented using OmpSs v.\,1.99.0,
calling the kernels directly using {\tt \#pragma omp task}
annotations with the respective dependencies, and
the runtime parameters
\begin{quote}
    \tt --disable-yield --schedule=socket --cores-per-socket=16 \\--num-sockets=4
\end{quote}
\noindent The scaling and efficiency relative to QuickSched are 
shown in \fig{QRResults}.
The difference in timings is the result of the different
task scheduling policies, as well as a smaller lag between the
individual tasks, as shown in \fig{QRTasks},
which shows the assignment of the different tasks to cores for the
64-core run.
The most visible difference between both schedulers is that
the DGEQRF tasks (in red) are scheduled as soon as they
become available in QuickSched, thus preventing bottlenecks
near the end of the computation.

@@

The Barnes-Hut tree-code \cite{ref:Barnes1986}
is an algorithm to approximate the
solution of an $N$-body problem, i.e.~computing all the
pairwise interactions between a set of $N$ particles,
in \oh{N\log N} operations, as opposed to the \oh{N^2}
naive direct computation.
The algorithm is based on a recursive octree decomposition:
Starting from a cubic cell containing all the particles,
the cell is recursively bisected along all three spatial dimensions,
resulting in eight sub-cells, until the number of particles
per cell is smaller than some limit $n_\mathsf{max}$.
The particle interactions can then be formulated recursively:
Given a particle and a set of particles in a cell,
if the particle and cell
are sufficiently well separated, the particle-cell interactions
are approximated by interacting the particle with the cell's
center of mass.
If the particle and the cell are too close, and the cell
has sub-cells, i.e.~it contained more than $n_\mathsf{max}$
particles and was split in the recursive octree decomposition,
then the particle is interacted with each of the sub-cells
recursively.
Finally, if the cell is not split, i.e.~it is a leaf cell
in the octree, then the particle is interacted with all
particles in the cell, except for the particle itself if
it is in the same cell.
This operation is performed for each particle, starting
with the root-level cell containing all the particles.

In our implementation, the particle data is sorted hierarchically,
following the octree structure.
Unlike in many codes, where the leaves store an array of
pointers to the underlying particles, which are not necessarily
contiguous in memory, the cells, at all
levels, store only a pointer to the first of their own particles,
and the total number of particles.
The current approach, illustrated in \fig{CellParts} is not
only more compact, it also allows a direct and more cache-efficient access
to the list of particles for any inner node of the tree.
The cost of sorting the particles, with a recursive
partitioning similar to QuickSort \cite{ref:Hoare1962},
is in \oh{N\log N}.

@@

This Barnes-Hut tree-code was used to approximate the gravitational
N-Body problem for 1\,000\,000 particles with uniformly random coordinates
in $[0,1]^3$.
The parameters $n_\mathsf{max}=100$ and $n_\mathsf{task}=5000$
were used to generate the tasks.
Using the above scheme generated 97\,553 tasks, of which
512 self-interaction tasks, 5\,068 particle-particle interaction
task, and 32\,768 particle-cell interaction tasks.
A total of 43\,416 locks on 37\,449 resources were generated.

@@

For comparison, the same computations were run using the popular
astrophysics simulation software Gadget-2 \cite{ref:Springel2005},
using a traditional Barnes-Hut implementation based on octrees
and distributed-memory parallelism based on domain decompositions
and MPI \cite{ref:Snir1998}.
To achieve the same accuracy, an opening angle of 0.5 was used.
On a single core, the task-based tree traversal is already 1.9$\times$
faster than Gadget-2, due to the cache efficiency of the task-based
computations, which, by design, maximize the amount of computation
per memory access.
At 59 cores, where Gadget-2 performs best, the task-based tree traversal is
2.51$\times$ faster, and at the full 64 cores it is 4$\times$ faster,
due to the better strong scaling of the task-based approach as opposed
to the MPI-based parallelism in Gadget-2.

@@

% Bibliography
\bibliographystyle{elsarticle-num}
\bibliography{quicksched}

@@

The following loops mirror the task generation described in
Algorithm~2 of \cite{ref:Buttari2009}.
For each level {\tt k} (line~10), a DGEQRF task is created
for tile $(k,k)$ (lines~13--14).
A lock is added for the newly created task on the
resource associated with the $(k,k)$th tile (line~15).
If a task exists at that position at the previous level
(line~16), a dependency is added between the old task and
the new (line~17), and the new task is stored in {\tt tid}
(line~18).
The remaining tasks are generated in the same way, with
their respective locks and dependencies.

>>>>

\documentclass[fleqn,10pt]{wlpeerj}

@@

% Title portion
\title{QuickSched: Task-based parallelism with dependencies and conflicts}
\author[1,3]{Pedro Gonnet}
\author[1]{Aidan B. G. Chalk}
\author[2]{Matthieu Schaller}
\affil[1]{School of Engineering and Computing Sciences, Durham University, United Kingdom.}
\affil[2]{Institute for Computational Cosmology, Durham University, United Kingdom.}
\affil[3]{Google Switzerland GmbH, Z\"urich, Switzerland.}

\keywords{task-based parallelism, shared-memory parallelism, high performance computing, parallel computing}

@@

\flushbottom
\maketitle
\thispagestyle{empty}

@@

Once modelled in such a way, the computation is somewhat trivial
to parallelize:
given a set of inter-dependent tasks and a set of computational
threads, each thread repeatedly selects a task with no
unsatisfied dependencies from the DAG and executes it.
If no tasks are available, the thread waits until any other
thread finishes executing a task, thus potentially releasing
new tasks, or until all tasks in the DAG have been executed.
Note that although the parallel execution
itself is trivial, it does not always guaranteed to be efficient.
Several factors may limit the maximum degree of parallelism, e.g.~the
structure of the task dependency DAG itself, or the order in which
available tasks are executed.

\fig{Tasks} shows such a DAG for a set of tasks.
The arrows indicate the direction of the dependency, i.e.~an
arrow from task $A$ to task $B$ indicates that task $B$ depends
on task $A$.
In a parallel setting, tasks $A$, $G$, and $J$ can be
executed concurrently.
Once task $G$ has completed, tasks $F$ and $H$ become available
and can be executed by any other computational thread.

@@

One of the first implementations of a task-based parallel programming
systems is Cilk \citep{ref:Blumofe1995}, an extension to the C
programming language which allows function calls to be ``spawned''
as new tasks.
Dependencies are enforced by the {\tt sync} keyword, which
forces a thread to wait for all the tasks that it spawned
to complete.

In SMP superscalar \citep{ref:Perez2008}, StarPU \citep{ref:Augonnet2011},
QUARK \citep{ref:Yarkhan2011}, and KAAPI \citep{ref:Gautier2007}
the programmer specifies
what shared data each task will access, and how that data will
be accessed, e.g.~read, write, or read-write access.
The dependencies between tasks are then generated
automatically by the runtime system, assuming that the
data must be accessed and updated in the order in which
the tasks are generated.
StarPU also provides an interface for specifying additional
dependencies explicitly.
Intel's Threading Building Blocks (TBB)
\citep{ref:Reinders2010}
provide task-based parallelism using C++ templates in which
dependencies are handled either by explicitly waiting
for spawned tasks, or by explicitly manipulating
task reference counters.

Finally, the very popular OpenMP standard provides some basic support
for spawning tasks, similar to Cilk, as of version 3.0
\citep{ref:OpenMP2008}.
OmpSs \citep{ref:Duran2011} extends this scheme with automatic
dependency generation as in SMP superscalar, of which it
is a direct descendant, along with
the ability to explicitly wait on certain tasks.

In all of these systems, the tasks are only aware of a single
type of relationship between each other, i.e. dependencies, which
specify a strict ordering between two tasks.
In many cases, however, the task ordering need not necessarily
be this strict.
Consider the case of two tasks that update some shared resource
in an order-independent way, e.g. when accumulating a result in
a shared variable, or exclusively writing to an output file.
In order to avoid concurrent access to that resource, it is
imperative that the execution of both tasks does not overlap,
yet the order in which the tasks are executed is irrelevant.
In the following, such a relationship will be referred to
as a {\em conflict} between two tasks.
\fig{TaskConflicts} shows a task graph extended by conflicting tasks
joined by thick dashed lines.
None of tasks $F$, $H$, and $I$ can be executed concurrently,
i.e. they must be serialized, yet in no particular order.

@@

From a programmer's perspective, there are two main paradigms for generating
task dependencies:
\begin{itemize}
  \item Implicitly via spawning and waiting, e.g.~as is done in Cilk
    \citep{ref:Blumofe1995}, or
  \item Automatic extraction from data dependencies, e.g.~as is done in OmpSs
    \citep{ref:Duran2011}.
\end{itemize}

@@

In the examples presented herein, we have chosen our task decomposition
and granularity such that
\begin{itemize}
  \item Each task maximizes the ratio of computation to data required,
  \item The resources required for each task fit comfortably in the
    lowest-level caches of the underlying system.
\end{itemize}
\noindent The first critera is biased towards bigger tasks, while the
second limits their size.
The parameters controlling the size of the tasks in the examples, 
i.e.~the tile size in the QR decomposition and the limits $n_\mathsf{max}$
and $n_\mathsf{task}$ were determined empirically and only optimized
to the closest power of two or rough power of ten, respectively.

@@

The QuickSched task scheduler consists of four main
objects types: {\em task}, {\em resource}, {\em scheduler},
and {\em queue}.
The task and resource objects are used
to model the computation, i.e. the work that is to be done
and the data on which it will be done, respectively.
The scheduler and queue objects manage
how the work is done, i.e. which tasks get scheduled
where and when, respectively.

@@

Finally, {\tt cost} and {\tt weight} are measures
for the relative computational cost of this task, and the
relative cost of the critical path following the
dependencies of this task, respectively, i.e.~the task's cost
plus the maximum dependent task's weight (see \fig{TaskWeight}).
The task cost can be either a rough estimate provided by the user,
or the actual cost of the same task last time it was executed.
The task weights are computed by traversing
the tasks DAG in reverse topological order following their dependencies,
e.g.~as per \cite{ref:Kahn1962}, and computing each task's weight, e.g.
\begin{equation*}
  \mbox{weight}_i = \mbox{cost}_i + \max_{j \in \mbox{\small unlocks}_i}\left\{\mbox{weight}_j\right\}.
\end{equation*}
\noindent where $\mbox{weight}_i$ and $\mbox{cost}_i$ are the
weight and cost of the $i$th task, respectively, and 
$\mbox{unlocks}_i$ is the set of tasks that the $i$th task
unlocks.

@@

Note also that although protecting the entire queue with a mutex
is not particularly scalable, and several authors, e.g.~REFS,
have presented concurrent data structures that avoid this type
of locking.
However, since we normally use one queue per computational thread,
contention will only happens due to work-stealing, i.e.~when
another idle computational thread tries to poach tasks.
Since this happens only rarely, we opt for the simpler locking approach.

@@

The tasks are executed as follows:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
void qsched_run(qsched *s, void (*fun)(int, void *)) {
  qsched_start(s);
  #pragma omp parallel
  {
    int qid = omp_get_thread_num() % s->nr_queues;
    struct task *t;
    while ((t = qsched_gettask(s, qid)) != NULL) {
      fun(t->type, t->data);
      qsched_done(s, t);
    }
  }
}
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where {\tt qsched\_start} initializes the tasks and
fills the queues (line~1).
For simplicity, OpenMP \citep{ref:Dagum1998}, which is available
for most compilers, is used to create a parallel section
in which the code between lines~4 and~11 is executed
concurrently.
A version using {\tt pthreads} \citep{ref:Pthreads1995}
directly\footnote{In most environments, OpenMP is implemented
on top of {\tt pthreads}, e.g. gcc's libgomp.} is also available.
The parallel section consists of a loop (lines~7--10) in
which a task is acquired via {\tt qsched\_gettask}
and its type and data are passed to a user-supplied
{\em execution function} {\tt fun}.
Once the task has been executed, it is returned to the
scheduler via the function {\tt qsched\_done}, i.e.~to
unlock its resources and unlock dependent tasks.
The loop terminates when the scheduler runs out of tasks,
i.e.~when {\tt qsched\_gettask} returns {\tt NULL}, and
the function exits once all the threads have exited their
loops.

@@

The function {\tt qsched\_gettask} fetches a task from
one of the queues:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
struct task *qsched_gettask(qsched *s, int qid) {
  struct task *res = NULL;
  int k;
  while (s->waiting) {
    if ((res = queue_get(s->queues[qid])) == NULL) {
      loop over all other queues in random order with index k
        if ((res = queue_get(s->queues[k])) != NULL)
          break;
    } else
      break;
  }
  if (res != NULL && s->reown) {
    for (k = 0; k < res->nr_locks; k++)
      res->locks[k]->owner = qid;
    for (k = 0; k < res->nr_uses; k++)
      res->uses[k]->owner = qid;
  }
  return res;
}
    \end{lstlisting}
\end{minipage}\end{center}
\noindent where the parameter {\tt qid} is the index of the
preferred queue.
If the queue is empty, or all of the tasks in that queue had
unresolved conflicts, the scheduler uses a variant of the
{\em work stealing} described in \cite{ref:Blumofe1999},
i.e.~it loops over all other queues
in a random order (line~6) and tries to get a task from them
(line~7).
If a task could be obtained from any queue and task re-owning
is enabled (line~12),
the resources it locks and uses are marked as now being owned
by the preferred queue (lines~13--16).
Finally, the task, or {\tt NULL} if no task could be obtained,
is returned.

@@

This section presents two test cases that show
how QuickSched can be used in real-world applications, and
provides benchmarks to assess its efficiency and scalability.
The first test is the tiled QR decomposition originally
described in \cite{ref:Buttari2009}, which has been used as a benchmark
by other authors \citep{ref:Agullo2009b,ref:Badia2009,ref:Bosilca2012}.
This example only requires dependencies and is presented 
as a point of comparison to existing task-based parallel
programming infrastructures.
The second example is a Barnes-Hut tree-code, a problem
similar to the Fast Multipole Method described in both
\cite{ref:Ltaief2012} and \cite{ref:Agullo2013}.
This example shows how conflicts, modeled
via hierarchical resources, can be useful in modelling and executing
a problem efficiently.

@@

\cite{ref:Buttari2009} introduced the concept of using task-based
parallelism for tile-based algorithms in numerical linear algebra,
presenting parallel codes for the Cholesky, LU, and QR
decompositions.
These algorithms are now part of the PLASMA and MAGMA
libraries for parallel linear algebra \citep{ref:Agullo2009}.
The former uses the QUARK task scheduler, which was originally
designed for this specific task, while the latter currently uses
the StarPU task scheduler \citep{ref:Agullo2011}.

@@

\noindent and where the task names are the BLAS-like operation
performed on the given tiles.
Every task depends on the task at the same position and the
previous level, i.e.~the task $(i,j,k)$ always depends on
$(i,j,k-1)$ for $k>1$.
Each task also modifies its own tile $(i,j)$, and the DTSQRF
task additionally modifies the lower triangular part of the $(j,j)$th tile.

Although the tile-based QR decomposition requires only dependencies,
i.e.~no additional conflicts are needed to avoid concurrent access to
the matrix tiles, we still model each tile as a separate resource
in QuickSched such that the scheduler can preferrentially assign
tasks using the same tiles to the same thread.
The resources/tiles are initially assigned to the queues in column-major
order, i.e.~the first $\lfloor n_\mathsf{tiles}/n_\mathsf{queues}\rfloor$
are assigned to the first queue, and so on.

The QR decomposition was computed for a $2048\times 2048$
random matrix using tiles of size $64\times 64$ floats using QuickSched
as described above.
The task costs were initialized to the asymptotic cost of the underlying
operations.
For this matrix, a total of 11\,440 tasks with 32\,240 dependencies
were generated.

For these tests, {\tt pthread} parallelism and resource re-owning
were used with one queue per core.
The QR decomposition was computed 10 times for each number of
cores, and the average thereof taken for the scaling and
efficiency results in \fig{QRResults}.
The timings are for {\tt qsched\_run}, including the cost of
{\tt qsched\_start}, which does not run in parallel.
Setting up the scheduler, tasks, and resources took, in all
cases, an average of 7.2\,ms, i.e.~at most 3\% of the total
computational cost.

The same decomposition was implemented using OmpSs v.\,1.99.0,
calling the kernels directly using {\tt \#pragma omp task}
annotations with the respective dependencies, and
the runtime parameters
\begin{quote}
  \tt --disable-yield --schedule=socket --cores-per-socket=16 \\--num-sockets=4
\end{quote}
\noindent Several different schedulers and parameterizations
were discussed with the authors of OmpSs and tested, with
the above settings produced the best results.

The scaling and efficiency relative to QuickSched are 
shown in \fig{QRResults}.
The difference in timings is the result of the different
task scheduling policies, as well as a smaller lag between the
individual tasks, as shown in \fig{QRTasks},
which shows the assignment of the different tasks to cores for the
64-core run.
The most visible difference between both schedulers is that
the DGEQRF tasks (in red) are scheduled as soon as they
become available in QuickSched, thus preventing bottlenecks
near the end of the computation.

@@

The Barnes-Hut tree-code \citep{ref:Barnes1986}
is an algorithm to approximate the
solution of an $N$-body problem, i.e.~computing all the
pairwise interactions between a set of $N$ particles,
in \oh{N\log N} operations, as opposed to the \oh{N^2}
naive direct computation.
The algorithm is based on a recursive octree decomposition:
Starting from a cubic cell containing all the particles,
the cell is recursively bisected along all three spatial dimensions,
resulting in eight sub-cells, until the number of particles
per cell is smaller than some limit $n_\mathsf{max}$.
The particle interactions can then be formulated recursively:
Given a particle and a set of particles in a cell,
if the particle and cell
are sufficiently well separated, the particle-cell interactions
are approximated by interacting the particle with the cell's
center of mass.
If the particle and the cell are too close, and the cell
has sub-cells, i.e.~it contained more than $n_\mathsf{max}$
particles and was split in the recursive octree decomposition,
then the particle is interacted with each of the sub-cells
recursively.
Finally, if the cell is not split, i.e.~it is a leaf cell
in the octree, then the particle is interacted with all
particles in the cell, except for the particle itself if
it is in the same cell.
This operation is performed for each particle, starting
with the root-level cell containing all the particles.

In our implementation, the particle data is sorted hierarchically,
following the octree structure.
Unlike in many codes, where the leaves store an array of
pointers to the underlying particles, which are not necessarily
contiguous in memory, the cells, at all
levels, store only a pointer to the first of their own particles,
and the total number of particles.
The current approach, illustrated in \fig{CellParts} is not
only more compact, it also allows a direct and more cache-efficient access
to the list of particles for any inner node of the tree.
The cost of sorting the particles, with a recursive
partitioning similar to QuickSort \citep{ref:Hoare1962},
is in \oh{N\log N}.

@@

This Barnes-Hut tree-code was used to approximate the gravitational
N-Body problem for 1\,000\,000 particles with uniformly random coordinates
in $[0,1]^3$.
The parameters $n_\mathsf{max}=100$ and $n_\mathsf{task}=5000$
were used to generate the tasks.
Using the above scheme generated 97\,553 tasks, of which
512 self-interaction tasks, 5\,068 particle-particle interaction
task, and 32\,768 particle-cell interaction tasks.
A total of 43\,416 locks on 37\,449 resources were generated.
Setting up the tasks took, on average XXX\,ms, i.e.~at most
XXX\% of the total computation time.
Storing the tasks, resources, and dependencies required XXX\,MB,
which is only XX\% of the XXX\,MB required to store the particle
data.

@@

For comparison, the same computations were run using the popular
astrophysics simulation software Gadget-2 \citep{ref:Springel2005},
using a traditional Barnes-Hut implementation based on octrees
and distributed-memory parallelism based on domain decompositions
and MPI \citep{ref:Snir1998}.
To achieve the same accuracy, an opening angle of 0.5 was used.
On a single core, the task-based tree traversal is already 1.9$\times$
faster than Gadget-2, due to the cache efficiency of the task-based
computations, which, by design, maximize the amount of computation
per memory access.
At 59 cores, where Gadget-2 performs best, the task-based tree traversal is
2.51$\times$ faster, and at the full 64 cores it is 4$\times$ faster,
due to the better strong scaling of the task-based approach as opposed
to the MPI-based parallelism in Gadget-2.

@@

% Bibliography
% \bibliographystyle{elsarticle-num}
\bibliography{quicksched}

@@

The following loops mirror the task generation described in
Algorithm~2 of \citep{ref:Buttari2009}.
For each level {\tt k} (line~10), a DGEQRF task is created
for tile $(k,k)$ (lines~13--14).
A lock is added for the newly created task on the
resource associated with the $(k,k)$th tile (line~15).
If a task exists at that position at the previous level
(line~16), a dependency is added between the old task and
the new (line~17), and the new task is stored in {\tt tid}
(line~18).
The remaining tasks are generated in the same way, with
their respective locks and dependencies.
